A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-06 paper/paper.tex
very briefly describe the models from Shulman's enriched indexed categories

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}

@@

Next, we introduce the contravariant ($\homlallXYtoZ \alpha R P$) and
covariant ($\homrallXYtoZ \alpha R P$) homs of sets, which are
different because we are in an orderd logic.
%
These are a kind of universally quantified
function type, where the universally quantified variable must occur
with the same variance in domain and codomain. In the contravariant
case, it occurs as the contravariant variable in both, and vice-versa
for the covariant case.
%
To highlight this, the notation for the contravariant dependence puts
the quantified variable on the \emph{left} of the triangle, as
contravariant variables occur to the left of the covariant variable,
and similarly the covariant hom has the quantified variable on the
right.
%
Then the
covariant variable of the contravariant hom set is the covariant
variable of the codomain and, and the contravariant variable of the
hom set is the \emph{covariant} variable of the domain, as the two
contravariances cancel. The covariant hom is dual.
%
Semantically, in ordinary category theory these are known as the
\emph{hom} of profunctors and are constructed using a combination of
an end and a function set.
%
The two connectives have similar introduction and elimination rules in
the form of $\lambda$ terms abstracting over both the object of the
category and the element of the set, and appropriate application
forms.
%
To keep with our invariant that the variable occurrences occur left to
right in the term syntax in a manner matching the context, we write
the covariant application in the usual order $\homrappXtoYatZ s t a$
where the function is on the left and the argument is on the right,
and the contravariant application in the flipped order.

@@

Finally, we briefly describe the connectives for the
``meta-logic''. These include the standard rules for Martin-L\"of type
theory with Pi/Sigma and extensional identity types, whose rules we
elide as they are standard \cite{emltt}. We use extensional identity
types so that the description of models is simpler, but there is no
obstacle to including models with intensional identity types.
%
The types we include are what we call \emph{universes} for the object
categorical logic: types of small categories and locally small
categories, functors, profunctors and natural transformations.

@@

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \varr{\cat C}{\cat D} \isaSet}

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \harr{\cat C}{\cat D} \isaSet}

@@

\section{Synthetic Category Theory in Ordered Higher-order Logic}

@@

\subsection{Categorical Models}

First, we can model the judgmental structure of the unary type theory
and predicate logic in certain\emph{virtual double categories}
\cite{leinster,crutwell-shulman}. We briefly recount the structure
present in a virtual double category, but see \citet{crutwell-shulman}
for a precise definition.
The following definition is
equivalent to the formulation as a ``normalized $T$-monoid'' in
\citet{crutwell-shulman}, and we direct the reader there for the full
description.
%

\begin{definition}[Virtual Double Category]
  A virtual double category $\mathcal V$ consists of
  \begin{enumerate}
  \item A category $V_o$ of ``objects and vertical arrows''
  \item A set ${\mathcal V}_h$ of ``horizontal arrows'' with source
    and target functions $s,t : {\mathcal V}_h \to {\mathcal V_o}^2$
  \item Sets of 2-cells of the following form:
% https://q.uiver.app/?q=WzAsNixbMCwwLCJDXzAiXSxbMCwyLCJEXzAiXSxbMiwwLCJDX3tufSJdLFsxLDAsIlxcY2RvdHMiXSxbMiwyLCJEXzEiXSxbMSwxLCJcXHBoaSJdLFszLDIsIlJfbiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9fX1dLFswLDEsIkYiLDJdLFsyLDQsIkciXSxbMSw0LCJTIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV0sWzAsMywiUl8wIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV1d
\[\begin{tikzcd}
	{C_0} & \cdots & {C_{n}} \\
	& \phi \\
	{D_0} && {D_1}
	\arrow["{R_n}", "\shortmid"{marking}, from=1-2, to=1-3]
	\arrow["F"', from=1-1, to=3-1]
	\arrow["G", from=1-3, to=3-3]
	\arrow["S"', "\shortmid"{marking}, from=3-1, to=3-3]
	\arrow["{R_0}", "\shortmid"{marking}, from=1-1, to=1-2]
\end{tikzcd}\]
  \end{enumerate}
  We say a virtual double category has \emph{split restrictions} when
  for any horizontal arrow $R : C \pto D$ and vertical arrows $f : C'
  \to C$ and $g : D' \to D$ there is a chosen profunctor $R(f,g) : C'
  \pto D'$ with a cartesian 2-cell to $R$ framed by $f,g$ and these
  chosen cartesian lifts are functorial in $f,g$.  A functor of
  virtual double categories preserves split restrictions if it
  preserves these specified cartesian lifts.  
\end{definition}

Each component of this definition has a direct correspondence to a
syntactic structure in \ohol{}. The category $\mathcal V_o$ models the
category and functor judgments. The set $\mathcal V_h$ models the
profunctor judgment. A composable string $R_0 \cdots R_n$ models the
profunctor contexts. The 2-cells correspond to the natural
transformation judgment where we have taken the restriction $S(F,G)$
of the codomain.

To model the dependent type theory and fact that the \ohol{} judgments
are all under a dependent type theory context, we use a variation on
Lawvere's notion of \emph{hyperdoctrine} for modeling predicate logic:
\begin{definition}[Hyperdoctrine of Virtual Double Categories with Restrictions]
  A hyperdoctrine of virtual double categories with restrictions is a
  pair of a category with families $\mathcal C$ and a functor $L :
  \mathcal C^o \to \text{vDbl}_r$ to the category of virtual double
  categories with split restrictions and restriction-preserving
  functors.
\end{definition}
Categories with families $\mathcal C$ model dependent type theory
\citep{cwf} and for each semantic context $\Gamma$, $L(\Gamma)$ models
the \ohol{} judgments in context $\Gamma$, with the functoriality
modeling the fact that all of these judgments admit a well-behaved
action of substitution.

@@

\subsection{Concrete Models}

Now that we have a category-theoretic notion of model, we give some
model construction theorems that can be used to justify our intuitive
notion of semantics in (enriched, internal) category theory.

First, we can extend any model of \ohol{} to a model of \ohol{}
parameterized by dependent type theory as follows:
\begin{construction}
  Given a virtual double category with restrictions $\mathcal V$, we
  can construct a functor $\mathcal V^{-}: \Set \to \text{vDbl}_r$ by
  defining the objects of $\mathcal V^\Gamma$ to be functions $\Gamma
  \to \mathcal V_o$, and similarly for morphisms and 2-cells with all
  operations given pointwise.
\end{construction}
% NOTE: this should be some kind of general theorem right?
% Sketch:
% 1. VDCrs are a finite limit theory
% 2. A hyperdoctrine L : C -> VDCr is the same as a VDCr in Pr(C)
% 3. The Yoneda embedding preserves finite limits

TODO: find a reference that proves things have all the UMPs we want
lol.

>>>>

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\MonCat}{\textrm{MonCat}}
\newcommand{\SymMonCat}{\textrm{SymMonCat}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}

@@

Next, we introduce the contravariant ($\homlallXYtoZ \alpha R P$) and
covariant ($\homrallXYtoZ \alpha R P$) homs of sets, which are
different because we are in an orderd logic.
%
These are a kind of universally quantified
function type, where the universally quantified variable must occur
with the same variance in domain and codomain. In the contravariant
case, it occurs as the contravariant variable in both, and vice-versa
for the covariant case.
%
To highlight this, the notation for the contravariant dependence puts
the quantified variable on the \emph{left} of the triangle, as
contravariant variables occur to the left of the covariant variable,
and similarly the covariant hom has the quantified variable on the
right.
%
Then the
covariant variable of the contravariant hom set is the covariant
variable of the codomain and, and the contravariant variable of the
hom set is the \emph{covariant} variable of the domain, as the two
contravariances cancel. The covariant hom is dual.
%
Semantically, in ordinary category theory these are known as the
\emph{hom} of profunctors and are constructed using a combination of
an end and a function set.
%
The two connectives have similar introduction and elimination rules in
the form of $\lambda$ terms abstracting over both the object of the
category and the element of the set, and appropriate application
forms.
%
To keep with our invariant that the variable occurrences occur left to
right in the term syntax in a manner matching the context, we write
the covariant application in the usual order $\homrappXtoYatZ s t a$
where the function is on the left and the argument is on the right,
and the contravariant application in the flipped order.
%
We also write the instantiating object as a superscript to
de-emphasize it, as in practice it can often be inferred.

@@

Finally, we briefly describe the connectives for the
``meta-logic''. These include the standard rules for Martin-L\"of type
theory with Pi/Sigma and extensional identity types, whose rules we
elide as they are standard \cite{emltt}. We use extensional identity
types so that the description of models is simpler, but there is no
obstacle to including models with intensional identity types.
%
The types we include are what we call \emph{universes} for the object
categorical logic: types of small categories and locally small
categories, functors, profunctors and natural transformations.
%
The rule for the types of small categories and categories are very
similar: any definable category defines an element of type $\Cats$,
and any element of that type can be reflected back into a
category. The only difference for $\smallCats$ is that the categories
involved additionally satisfy $\cat C \isSmall$. Again we elide the
standard $\beta\eta$ principles.
%
Since every small category $\cat C \isSmall$ is a category $\cat C
\isaCat$, there is a definable inclusion function from $\smallCats$ to
$\Cat$ and the $\beta\eta$ properties ensure that this is a
monomorphism.

Next, we have the types of all functors and profunctors between any
two fixed categories.
%
The introduction and elimination forms are those for unary and binary
function types respectively.
%
Finally we include a type $\pendallXdotY {\alpha:\cat C} P$ which we
call the set of ``natural elements'' of $P$.
%
The name comes from the case that $P$ is of the form
$\punitinXfromYtoZ {}{F(\alpha)}{G(\alpha)}$ in which case the type
$\pendallXdotY {\alpha:\cat C} \punitinXfromYtoZ
{}{F(\alpha)}{G(\alpha)}$ can be interpreted as the set of all natural
transformations from $F$ to $G$.
%
More generally this is modeled as an end, and matching our notation
for hom types we use a universal quantifier to notate this.

@@

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \varr{\cat C}{\cat D} \isaTy}

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \harr{\cat C}{\cat D} \isaTy}

@@

\section{Synthetic Category Theory in VETT}

@@

First, we can model the judgmental structure of the unary type theory
and predicate logic in certain\emph{virtual equipments}
\cite{leinster,crutwell-shulman}, i.e., virtual double categories with
restrictions. We briefly recount the structure present in a virtual
double category, but see \citet{crutwell-shulman} for a precise
definition of the composition rules for 2-cells and functor of virtual
double categories.
%
Our notion here differs from \citet{crutwell-shulman} in several ways
to more closely match our judgmental structure: (1) we require that
restrictions are given as coherent structure rather than mere
existence (2) we include a notion of ``small'' object and (3) we don't
yet require the existence of unit horizontal arrows, as these are
modeled by a connective rather than judgmental structure.

\begin{definition}[Virtual Equipment]
  A virtual equipment $\mathcal V$ consists of
  \begin{enumerate}
  \item A category $V_o$ of ``objects and vertical arrows''
  \item A chosen subset of the objects $V_s \subseteq V_o$ of ``small objects''
  \item A set ${\mathcal V}_h$ of ``horizontal arrows'' with source
    and target functions $s,t : {\mathcal V}_h \to {\mathcal V_o}^2$
  \item Sets of 2-cells of the following form:
% https://q.uiver.app/?q=WzAsNixbMCwwLCJDXzAiXSxbMCwyLCJEXzAiXSxbMiwwLCJDX3tufSJdLFsxLDAsIlxcY2RvdHMiXSxbMiwyLCJEXzEiXSxbMSwxLCJcXHBoaSJdLFszLDIsIlJfbiIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9fX1dLFswLDEsIkYiLDJdLFsyLDQsIkciXSxbMSw0LCJTIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV0sWzAsMywiUl8wIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn19fV1d
\[\begin{tikzcd}
	{C_0} & \cdots & {C_{n}} \\
	& \phi \\
	{D_0} && {D_1}
	\arrow["{R_n}", "\shortmid"{marking}, from=1-2, to=1-3]
	\arrow["f"', from=1-1, to=3-1]
	\arrow["g", from=1-3, to=3-3]
	\arrow["S"', "\shortmid"{marking}, from=3-1, to=3-3]
	\arrow["{R_0}", "\shortmid"{marking}, from=1-1, to=1-2]
\end{tikzcd}\]
    With appropriate ``multi-categorical'' notions of identity and
    composition.

    We say that the 2-cell $\phi$ has $S$ as codomain, the sequence
    $R_0 \ldots R_n$ as domain and call $f$ and $g$ the left and right
    ``frames'', or that $\phi$ is framed by $f$ and $g$.
  \end{enumerate}
  We say a virtual double category has \emph{split restrictions} when
  for any horizontal arrow $R : C \pto D$ and vertical arrows $f : C'
  \to C$ and $g : D' \to D$ there is a chosen horizontal arrow $R(f,g)
  : C' \pto D'$ with a cartesian 2-cell to $R$ framed by $f,g$ and
  these chosen cartesian lifts are functorial in $f,g$.

  A functor of virtual equipments is a functor of the virtual double
  categories that additionally preserves the restrictions and
  smallness of objects.
\end{definition}

In the presence of restrictions, every 2-cell can be represented as a
``globular'' 2-cell where the left and right frame are identities. For
example the 2-cell $\phi$ above can be represented as one with the
same domain but whose codomain is $S(f,g)$. This property is crucial
for the completeness of our semantics as we only include a syntax for these globular terms 
%
Each component of this definition has a direct correspondence to a
syntactic structure in \ohol{}. The objects of $\mathcal V_o$ models
the category judgment and the morphisms model the functor
judgment. The set $\mathcal V_h$ models the profunctor judgment. A
composable string $R_0 \cdots R_n$ models the profunctor contexts. The
2-cells correspond to the natural transformation judgment where we
have taken the restriction $S(F,G)$ of the codomain.

To model the dependent type theory and fact that the categorical
judgments are all typed under a context $\Gamma$ with an action of
substitution, we use a variation on Lawvere's notion of
\emph{hyperdoctrine} for modeling predicate logic:
\begin{definition}[Hyperdoctrine of Virtual Equipments]
  A hyperdoctrine of virtual equipments is a pair of a category with
  families $\mathcal C$ and a functor $V^{(-)} : \mathcal C^o \to
  \text{vEq}$ to the category of virtual equipments and functors.
\end{definition}
Categories with families $\mathcal C$ model dependent type theory
\citep{cwf} and for each semantic context $\Gamma$, $V^\Gamma$ models
the VETT judgments in context $\Gamma$, with the functoriality
modeling the fact that all of these judgments admit a well-behaved
action of substitution.

@@

\subsection{Examples of Models}

Now that we have a category-theoretic notion of model, we give some
model construction theorems that can be used to justify our intuitive
notion of semantics in (enriched, internal, indexed) category theory.

First, we can extend any virtual equipment to a hyperdoctrine of
virtual equipments

First, we can extend any model of \ohol{} to a model of \ohol{}
parameterized by dependent type theory as follows:
\begin{construction}
  Given a virtual equipment $\mathcal V$, we
  can construct a functor $\mathcal V^{-}: \Set \to \text{vDbl}_r$ by
  defining the objects of $\mathcal V^\Gamma$ to be functions $\Gamma
  \to \mathcal V_o$, and similarly for morphisms and 2-cells with all
  operations given pointwise. 
\end{construction}
% NOTE: this should be some kind of general theorem right?
% Sketch:
% 1. VDCrs are a finite limit theory
% 2. A hyperdoctrine L : C -> VDCr is the same as a VDCr in Pr(C)
% 3. The Yoneda embedding preserves finite limits

%% Semantic Qs:
%
%% (1) The virtual equipment of locally small categories is a model of everything.
%% (2) Are Shulman's enriched indexed categories strict enough?

Then to define a model of VETT with a collection of connectives is is
sufficient to construct a virtual equipment with the universal
properties corresponding to those connectives.
%
\begin{construction}
  Fix a cardinal $\kappa$. The virtual equipment $\Cat_{\kappa}$ is
  defined to have as objects locally $\kappa$-small categories, small
  objects as $\kappa$-small categories, vertical morphisms as functors,
  horizontal arrows as functors $\cat C^o \times \cat D \to \kappa Set$ and
  2-cells as morphisms of profunctors. Restriction of profunctors is
  given by composition, which is strictly associative and unital.
  $\Cat_{U}$ has objects satisfying the universal properties of all
  connectives in VETT.
\end{construction}

More generally, categories internal to, enriched in and/or indexed by
sufficiently nice categories define a virtual equipment that model
most of the connectives of our type theory. We highlight one example
from the literature that is highly general: Shulman's enriched indexed
categories. Shulman's construction defines a virtual double categories
of large and small $\mathcal V$-categories for any pseudofunctor
$\mathcal V : S^o \to \MonCat$ where $S$ is a category with finite
products. He gives examples that show that this internal, enriched and
indexed categories for suitable choices of $\mathcal V$.
%
This is slightly weaker then what we require: to have
\emph{split} restrictions, we need that $\mathcal V$ be a
\emph{strict} functor, not merely a pseudo-functor.
%
This is analogous to the situation for dependent type theory where
substitution is strictly associative but in models is typically
modeled by pullback which is only associative up to unique
isomorphism. In fact, one of the main examples is only a
pseudo-functor for precisely this reason that the action on morphisms
is given by pullback.
%
We expect this difference can be handled in the same way: by a general
strictification theorem or by moving to a univalent setting.

\begin{construction}[(\citet{shulman-enriched-indexed})]
  Given any functor $\mathcal V : S^o \to \SymMonCat$ such that $S$
  and $\mathcal V$ have sufficiently well-behaved (indexed)
  $\kappa$-products, then there is a virtual equipment $\mathcal
  V-\textrm{Cat}$ whose objects are locally $\kappa$-small $\mathcal
  V$-categories, small objects are $\kappa$-small $\mathcal
  V$-categories etc. This virtual equipment has been shown in
  loc. cit. to have objects satisfying all of the universal properties
  used in models of VETT except the graph of a profunctor.
\end{construction}
